Erosion of somatic tissue identity with loss of the X-linked intellectual disability factor KDM5C

### 2024-08-20 submission/manuscript.Rmd
last 3 discussion paragraphs

Our analysis of KDM5C ChIP-seq data provided novel insights into direct and indirect mechanisms by which different classes of germline genes can be misexpressed in _Kdm5c_-KO cells. Transcription of germline genes in the mature _Kdm5c_-KO brain is likely due to loss of KDM5c-mediated repression during embryogenesis, given that KDM5C was bound to their promoters in EpiLCs, is not recruited to germline genes in neurons, and germline transcription is not repressed when KDM5C in reintroduced in neurons[@scandagliaLossKdm5cCauses2017]. It is unclear how KDM5C is recruited to germline genes, given that KDM5C itself does not contain domains for sequence-specific binding. In HeLa cells and ESCs[@tahilianiHistoneH3K4Demethylase2007; @guptaGenomewideScreenReveals2023], KDM5C assocaiates with members of the polycomb repressive complex 1.6 (PRC1.6), which is recruited to germline gene promoters through E2F6/DP1 and MGA/MAX heterodimers[@endohPCGF6PRC1SuppressesPremature2017; @stielowMGAL3MBTL2E2F62018]. While MAX and E2F6 motifs were enriched at KDM5C-bound germline promoters in EpiLCs, only about one-third of promtoers contained their consesus sequence. Additionally, germline genes like _Stra8_ can become active in _Kdm5c_-KO cells independent of KDM5C binding to their promoters. _Stra8_ is a meiotic transcription factor that is typically turned on in the presence of retinoic acid signaling and DAZL[@endoPeriodicRetinoicAcid2015; @bowlesRetinoidSignalingDetermines2006], which is a direct KDM5C target. Germline genes unbound by KDM5C were primarily late-stage spermatogenesis genes and significantly enriched for RFX transcription factors that promote the transcription of cilia and flagellar genes[@swobodaRFXTypeTranscriptionFactor2000; @ashiqueRfx4TranscriptionFactor2009]. One of the top RFX members enriched included RFX2, a central regulator of spermiogenesis[@kistlerRFX2MajorTranscriptional2015; @wuTranscriptionFactorRFX22016] whose mRNA is also expressed in _Kdm5c_-KO EpiLCs. Altogether, these data suggest that once germline genes are activated due to loss of direct KDM5C repression, they can activate other ectopic germline programs downstream that loosely mimic germline development.

Unlike the previously characterized germline gene suppressors[@karimiDNAMethylationSETDB12011; @hackettPromoterDNAMethylation2012; @auclairEHMT2DirectsDNA2016; @endohPCGF6PRC1SuppressesPremature2017; @tatsumiDNMTsSETDB1Function2018; @liuPolycombGroupProtein2020; @mochizukiRepressionGermlineGenes2021], KDM5C removes an active histone mark, histone 3 lysine 4 di- and trimethylation (H3K4me2/3)[@iwaseXLinkedMentalRetardation2007]. However, KDM5C's catalytic activity may not be required for germline gene silencing, as it was recently found to be dispensible for repressing _Dazl_ in ESCs [@guptaGenomewideScreenReveals2023]. Emerging work indicate many histone-modifying enzymes have non-cataltyic functions that can influnce gene expression at least as potently as their catalytic roles[@aubertUnexpectedNoncatalyticRoles2019; @morganReevaluatingRolesHistonemodifying2020]. While dispensible in ESCs, KDM5C's catalytic activity may be necessary in EpiLCs, as H3K4me3 impedes _de novo_ CpGme placement[@otaniStructuralBasisRecognition2009; @guoStructuralInsightAutoinhibition2015] and CpGme is significantly eroded at at least two germline promoters in the adult _Kdm5c_-KO hippocampus[@scandagliaLossKdm5cCauses2017]. Indeed, we found _Kdm5c_-KO EpiLCs have impaired CpGme accumulation at germline gene CpG islands, indicating H3K4me3 removal is necessary at this developmental stage. Thus, the shift in the mechanism of KDM5C-mediated repression during the transition from na&#239;ve to primed pluripotency demonstrates the cataltyic activity of histone-modifying enzymes can become necessary within different cellular contexts.

- DNAme and CpG islands
	- Loss of many chromatin regulators affects CpGme (Multiple, non-redundant silencing mechansisms) - seems to be the crux point of germline gene silencing
	- Germline genes are methylated at promoter CGIs, which are typically not methylated in other gene types
	- Because CGIs are typically resistant to CpGme (accurate?), germline CGIs may require a highly repressive histone landscape to recruit sufficient DNMTs to these loci
	- We found not all germline-enriched genes have CGIs, but still gain CpGme around TSS. Unclear what the recruitment mechanism is for these genes
	- Other studies on germline gene repressors have shown they are importnat for CGIme, unclear if they also participate in non-CGI TSS CpGme
	- CGIs highly determine KDM5C recruitment to germline gene promoters and KDM5C germline gene CGI methylation.
		- KDM5C is known to be enriched at CGIs (in neurons? are these methylated? or is its germline CGI function likely different from its somatic CGI function). 
	- Different classes of germline genes use distinct silencing mechanisms

The above work provides the mechanistic foundation for KDM5C's role in germline gene repression and offers novel insight into how dysregulation in _Kdm5c_-KO tissue identity changes over the course of development. However, the contribution of ectopic, tissue-specific genes towards _Kdm5c_-KO neurological impairments is still unknown. In addition to germline genes, we also found significant enrichment of muscle, liver, and even ovary-biased genes aberrantly expressed within the male _Kdm5c_-KO brain. Intriguingly, select liver and muscle-biased DEGs do have known roles within the brain, such as the liver-enriched lipid metabolism gene _Apolipoprotein C-I (Apoc1)_ [@roulandRoleApolipoproteinC12022] that is highly expressed in the _Kdm5c_-KO amygdala and hippocampus. _Apoc1_ overexpression in the mouse brain can impair learning and memory[@abildayevaHumanApolipoproteinCI2008] and is implicated in Alzheimer's disease in humans[@leducAPOECholesterolHomeostasis2010]. Thus, KDM5C may be crucial for neurodevelopment by fine-tuning the expression of tissue-erniched, dosage-sensitive genes like _Apoc1_. While the impact of ectopic germline transcripts upon neurodevelopment is currently unknown, some models of other chromatin-linked neurodevelopmental disorders also display impaired soma-germline demarcation[@bonefasSomatogermlineTransformationChromatinlinked2021; @velascoGermlineGenesHypomethylation2014; @waltonDnmt3bPrefersGerm2014; @schaeferControlCognitionAdaptive2009; @samacoCrhOprm1Mediate2012]. Like KDM5C, the chromatin regulators underlying these conditions - DNA methyltransferase 3b (DNMT3B), H3K9me1/2 methyltransferases G9A/GLP, methyl-CpG -binding protein 2 (MECP2) - primarily silence gene expression. Thus, KDM5C is among a growing cohort of chromatin-linked neurodevelopmental disorders that have a similar underlying phenotype of germline versus soma dysregulation, however further research is required to determine the impact of these germline genes and the extent to which this phenomenon occurs in humans.

>>>>

It is yet unknown how KDM5C is recruited to germline gene promoters, given that KDM5C itself does not contain domains for sequence-specific binding. In HeLa cells and ESCs[@tahilianiHistoneH3K4Demethylase2007; @guptaGenomewideScreenReveals2023], KDM5C assocaiates with members of the polycomb repressive complex 1.6 (PRC1.6), which is recruited to germline gene promoters through E2F6/DP1 and MGA/MAX heterodimers[@endohPCGF6PRC1SuppressesPremature2017; @stielowMGAL3MBTL2E2F62018]. While MAX and E2F6 motifs were enriched at KDM5C-bound germline promoters in EpiLCs, only about one-third of promoters contained their consesus sequence. Additionally, some germline genes can become active in _Kdm5c_-KO cells independent of KDM5C binding to their promoters. Germline genes unbound by KDM5C were primarily late-stage spermatogenesis genes and significantly enriched for RFX transcription factors, including RFX2, a central regulator of spermiogenesis[@kistlerRFX2MajorTranscriptional2015; @wuTranscriptionFactorRFX22016]. Another notable KDM5C-unbound DEG is _Stra8_, a meiotic transcription factor that is turned on in germ cells with retionic acid signaling and DAZL expression[@endoPeriodicRetinoicAcid2015; @bowlesRetinoidSignalingDetermines2006]. Together, this suggests once germline genes like _Dazl_ and _Rfx2_ are activated in _Kdm5c_-KO cells, they can then turn on other ectopic germline programs downstream and loosely mimic germ cell development.

Unlike the previously characterized germline gene suppressors that place repressive histone modifications[@karimiDNAMethylationSETDB12011; @hackettPromoterDNAMethylation2012; @auclairEHMT2DirectsDNA2016; @endohPCGF6PRC1SuppressesPremature2017; @tatsumiDNMTsSETDB1Function2018; @liuPolycombGroupProtein2020; @mochizukiRepressionGermlineGenes2021], KDM5C removes the active histone mark histone 3 lysine 4 di- and trimethylation (H3K4me2/3)[@iwaseXLinkedMentalRetardation2007]. Yet, KDM5C's catalytic activity may not be required for germline gene silencing, as it was recently found to be dispensible for repressing _Dazl_ in ESCs [@guptaGenomewideScreenReveals2023]. Emerging work indicates many histone-modifying enzymes have non-cataltyic functions that can influnce gene expression even more potently as their catalytic roles[@aubertUnexpectedNoncatalyticRoles2019; @morganReevaluatingRolesHistonemodifying2020]. Although dispensible in ESCs, KDM5C's catalytic activity could be necessary to silence germline genes in EpiLCs, as H3K4me3 impedes _de novo_ CpGme placement[@otaniStructuralBasisRecognition2009; @guoStructuralInsightAutoinhibition2015]. In support of this, CpGme is significantly eroded at at least two germline promoters in the adult _Kdm5c_-KO hippocampus[@scandagliaLossKdm5cCauses2017]. Additionally, we found _Kdm5c_-KOs accumulate H3K4me2/3 around the TSS of germline genes and impaired CpGme placement at CpG islands. KDM5C's shifting mode of germline gene repression as cells transition from na&#239;ve to primed pluripotency indicates chromatin regulators can employ both catalytic and non-catlytic repressive mechanisms at the same loci depending upon the developmental and cellular context.

- DNAme and CpG islands
	- Loss of many chromatin regulators affects CpGme (Multiple, non-redundant silencing mechansisms) - seems to be the crux point of germline gene silencing
	- Germline genes are methylated at promoter CGIs, which are typically not methylated for other types of genes
	- Because CGIs are typically resistant to CpGme (accurate?), germline CGIs may require a highly repressive histone landscape to recruit sufficient DNMTs to these loci
	- We found not all germline-enriched genes have CGIs, but many still gain CpGme around the TSS. Unclear what the repressive mechanism is for these genes
	- CGIs highly determine KDM5C recruitment to germline gene promoters.
			- KDM5C is known to be enriched at CGIs (in neurons? are these methylated? or is its germline CGI function likely different from its somatic CGI function). 
	- KDM5C loss impacts germline gene CGI methylation, but not really the non-CGI TSS methylation
		- Other studies on germline gene repressors have shown they are importnat for CGIme, unclear if they participate in non-CGI TSS CpGme
	- Altogether demonstrates different classes of germline genes use distinct silencing mechanisms

The above work provides the mechanistic foundation for KDM5C's role in germline gene repression and offers novel insight into how dysregulation in _Kdm5c_-KO tissue identity changes over the course of development. However, the contribution of these ectopic, tissue-specific genes towards _Kdm5c_-KO neurological impairments is still unknown. In addition to germline genes, we also identified significant enrichment of muscle, liver, and even ovary-biased transcripts within the male _Kdm5c_-KO brain. Intriguingly, select liver and muscle-biased DEGs do have known roles within the brain, such as the liver-enriched lipid metabolism gene _Apolipoprotein C-I (Apoc1)_ [@roulandRoleApolipoproteinC12022] that can impair learning and memory when overexpressed in the mouse brain[@abildayevaHumanApolipoproteinCI2008] and is implicated in Alzheimer's disease in humans[@leducAPOECholesterolHomeostasis2010]. Thus, KDM5C may be crucial for neurodevelopment by fine-tuning the expression of tissue-erniched, dosage-sensitive genes like _Apoc1_. Given germline genes have no known functions within the brain, their impact upon neruodevelopment is currently unknown. Ectopic germline transcripts have been observed in a variety of cancers[@nielsenEctopicExpressionTestis2016; @adebayobabatundeCancerTestisGenes2017] and can drive brain turmor formation in _Drosophila_[@janicEctopicExpressionGermline2010], indicating they may promote genome instability and cellular de-differentiation. Intriguingly, select models of other chromatin-linked neurodevelopmental disorders also display impaired soma-germline demarcation[@bonefasSomatogermlineTransformationChromatinlinked2021; @velascoGermlineGenesHypomethylation2014; @waltonDnmt3bPrefersGerm2014; @schaeferControlCognitionAdaptive2009; @samacoCrhOprm1Mediate2012]. Like KDM5C, the chromatin regulators underlying these conditions - DNA methyltransferase 3b (DNMT3B), H3K9me1/2 methyltransferases G9A/GLP, methyl-CpG -binding protein 2 (MECP2) - primarily silence gene expression. Thus, KDM5C is among a growing cohort of chromatin-linked neurodevelopmental disorders that have a similar underlying phenotype of germline versus soma dysregulation However, further research is required to determine the impact of these germline genes and the extent to which this phenomenon occurs in humans.
